---
title: "Wrangling Variables"
author: "Name: SOLUTIONS"
lang: en
format:
  pdf:
    documentclass: article
    fontsize: 12pt
    geometry:
      - margin=1in
    colorlinks: true
    number-sections: false
    header-includes: |
      \usepackage{xcolor}
      \usepackage{titlesec}
      \usepackage{fancyhdr}
      \usepackage{hyperref}
      \usepackage{graphicx}

      % ---- Brand colors, pulled from the slide theme (style.scss) ----
      % Kept at the same WCAG AA/AAA-checked hex values as the slides.
      \definecolor{harvardcrimson}{HTML}{A51C30}
      \definecolor{stattextmuted}{HTML}{595959}
      \definecolor{tealaccent}{HTML}{006677}
      \definecolor{slateaccent}{HTML}{3B4B61}
      \definecolor{emeraldaccent}{HTML}{1E6B3F}
      \definecolor{bodycharcoal}{HTML}{1E1E1E}
      \definecolor{ruleline}{HTML}{595959}

      % Section headings in crimson, matching the slide deck's heading color.
      % Weight + color together (not color alone) carry the emphasis.
      \titleformat{\section}{\color{harvardcrimson}\Large\bfseries}{\thesection}{0.6em}{}
      \titleformat{\subsection}{\color{harvardcrimson}\large\bfseries}{\thesubsection}{0.6em}{}

      % Recolor Quarto's built-in callout boxes to match the slide accent palette
      \definecolor{quarto-callout-color-note}{HTML}{006677}
      \definecolor{quarto-callout-color-tip}{HTML}{1E6B3F}
      \definecolor{quarto-callout-color-important}{HTML}{A51C30}
      \definecolor{quarto-callout-color-caution}{HTML}{3B4B61}

      % Footer to match the slide deck's footer link, plus the course logo
      % in the upper-right corner (same asset used in the slide theme).
      % Wrapped in \IfFileExists so the document still compiles even if
      % images/logo.png hasn't been added yet -- it just shows no logo.
      \newcommand{\brandlogo}{\IfFileExists{images/logo.png}{\includegraphics[height=0.5cm]{images/logo.png}}{}}
      \pagestyle{fancy}
      \fancyhf{}
      \renewcommand{\headrulewidth}{0pt}
      \fancyhead[R]{\brandlogo}
      \fancyfoot[C]{\small\textcolor{stattextmuted}{stat-100.com}}
      \fancyfoot[R]{\small\textcolor{stattextmuted}{\thepage}}

      % article's \maketitle uses the "plain" pagestyle for the title page,
      % which would otherwise hide the header/logo on page 1 -- redefine it
      % to match.
      \fancypagestyle{plain}{%
        \fancyhf{}
        \renewcommand{\headrulewidth}{0pt}
        \fancyhead[R]{\brandlogo}
        \fancyfoot[C]{\small\textcolor{stattextmuted}{stat-100.com}}
        \fancyfoot[R]{\small\textcolor{stattextmuted}{\thepage}}
      }

      \hypersetup{
        linkcolor=harvardcrimson,
        urlcolor=harvardcrimson,
        citecolor=harvardcrimson,
        bookmarksopen=true,
        pdfdisplaydoctitle=true
      }

      % Helper LaTeX command used below to build write-in space (a blank line)
      \newcommand{\answerline}{\par\vspace{0.55cm}\textcolor{ruleline}{\rule{\linewidth}{0.6pt}}\par}

      % A bordered blank box for students to write R code in (Part 2).
      % Pure core LaTeX (fbox + minipage) -- no extra package needed.
      % NOTE: a minipage containing only \vspace (no actual horizontal
      % material) collapses to near-zero width when framed with \fbox --
      % the \rule{\linewidth}{0pt} below is a zero-height rule that just
      % anchors the box to its full width before the vspace runs.
      \newcommand{\codebox}[1][3.2]{%
        \par\vspace{0.15cm}
        \noindent\fbox{%
          \begin{minipage}[t]{0.97\linewidth}
            \rule{\linewidth}{0pt}\vspace{#1cm}
          \end{minipage}%
        }
        \par\vspace{0.15cm}
      }
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(tibble)
options(width = 90, pillar.width = 90)

# A small, 10-row x 5-column data set students can work with by hand.
# Streaming-song data (not board games). One genre is deliberately left
# missing (NA) so the is.na() exercise has something real to find.
songs <- tibble::tribble(
  ~track_name,        ~genre,     ~release_year, ~duration_sec, ~streams_millions,
  "Blinding Lights",  "Pop",               2020,           200,                87,
  "Bad Guy",          "Pop",               2019,           194,                65,
  "Levitating",       "Pop",               2020,           203,                72,
  "Sunflower",        "Hip-Hop",           2018,           158,                58,
  "Perfect",          "Pop",               2017,           263,                45,
  "Old Town Road",    "Hip-Hop",           2019,           113,                90,
  "Shape of You",     "Pop",               2017,           233,                55,
  "Circles",          "Indie",             2019,           215,                30,
  "Ocean Eyes",       "Indie",             2015,           200,                12,
  "Heat Waves",       NA,                  2020,           238,                68
)

# A blank write-in placeholder: a run of underscores, used as plain text
# inside a normal printed tibble (not a LaTeX table), so there's nothing
# for LaTeX to choke on and no float that can drift to another page.
blank_col <- function(n, width = 16) rep(strrep("_", width), n)

# Part 2's data set: family movies, 10 rows x 5 columns. Genre is
# deliberately missing for one row, same as `songs`, so Part 2 can reuse
# the same is.na() idea.
movies <- tibble::tribble(
  ~title,               ~genre,       ~release_year, ~runtime_min, ~box_office_millions,
  "Coco",               "Animation",           2017,          105,                   72,
  "Moana",              "Animation",           2016,          107,                   84,
  "Up",                 "Animation",           2009,           96,                   68,
  "Zootopia",           "Family",              2016,          108,                   91,
  "Encanto",            "Animation",           2021,          102,                   58,
  "Toy Story 4",        "Family",              2019,          100,                   76,
  "Inside Out",         "Family",              2015,           95,                   65,
  "Incredibles 2",      "Adventure",           2018,          118,                   88,
  "Finding Dory",       "Adventure",           2016,           97,                   62,
  "Soul",               NA,                    2020,          100,                   45
)
```

# Part 1

## The Data

This is the data set you'll work with for every exercise in Part 1 of this handout:
`songs`, 10 rows and 5 columns.

```{r}
#| label: show-main
songs
```

\newpage

## Exercise 1 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  rename(
    music_genre = genre,
    stream_count_millions = streams_millions
  )
```

Show the changes made to the `songs` data frame below by writing on the values/variables and/or crossing rows and/or columns.

```{r}
#| label: show-rename
songs |>
  rename(
    music_genre = genre,
    stream_count_millions = streams_millions
  )
```


\newpage

## Exercise 2 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  select(track_name, release_year, streams_millions)
```

Show the changes made to the `songs` data frame below by writing on the values/variables and/or crossing rows and/or columns.


```{r}
#| label: show-select
songs |>
  select(track_name, release_year, streams_millions)
```

\newpage

## Exercise 3 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  filter(release_year >= 2019)
```

Shade the rows below that would remain after running this
code.

```{r}
#| label: show-filter
songs |>
  filter(release_year >= 2019)
```

How many rows would remain? 6

\newpage

## Exercise 4 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  filter(genre == "Pop" & streams_millions >= 60)
```

Shade the rows below that satisfy **both** conditions.

```{r}
#| label: show-filter2
songs |>
  filter(genre == "Pop")
```

Answer: 

```{r}
#| label: show-filter3
songs |>
  filter(genre == "Pop" & streams_millions >= 60)
```

\newpage

## Exercise 5 

Suppose we run:

```{r}
#| echo: true
#| eval: false
specific_years <- c(2017, 2020)

songs |>
  filter(release_year %in% specific_years)
```

Shade the rows below whose `release_year` is in
`specific_years`.

```{r}
#| label: show-in-operator
specific_years <- c(2017, 2020)

songs |>
  filter(release_year %in% specific_years)
```

\vspace{0.4cm}
How many rows would remain? 5

\newpage

## Exercise 6 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  filter(is.na(genre))
```

Shade the row(s) below that would remain after running the code above.

```{r}
#| label: show-isna
songs |>
  filter(is.na(genre))
```

\newpage

## Exercise 7 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  mutate(
    viral_status = if_else(
      streams_millions >= 50,
      "Viral hit",
      "Not yet viral"
    )
  )
```

Fill in only the new `viral_status` column below (write over
the blanks).

```{r}
#| label: show-mutate-ifelse
songs |>
  mutate(
    viral_status = if_else(
      streams_millions >= 50,
      "Viral hit",
      "Not yet viral"
    )
  )
```

\newpage

## Exercise 8 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  mutate(
    stream_tier = case_when(
      streams_millions < 20  ~ "Under 20M",
      streams_millions < 50  ~ "20M to 49M",
      streams_millions >= 50 ~ "50M or more"
    )
  )
```

Fill in only the new `stream_tier` column below (write over the blanks).

```{r}
#| label: show-mutate-casewhen
songs |>
  mutate(
    stream_tier = case_when(
      streams_millions < 20  ~ "Under 20M",
      streams_millions < 50  ~ "20M to 49M",
      streams_millions >= 50 ~ "50M or more"
    )
  )
```

\newpage

## Exercise 9 

Suppose we run:

```{r}
#| echo: true
#| eval: false
songs |>
  group_by(genre) |>
  summarize(
    med_duration_sec = median(duration_sec),
    count = n()
  )
```

Use Exercise 2's table to work
out the median duration and count of songs for each genre. The group labels
are already filled in — write your answers over the blanks. Remember: the
missing `genre` value forms its own group.

```{r}
#| label: show-groupby
songs |>
  group_by(genre) |>
  summarize(
    med_duration_sec = median(duration_sec),
    count = n()
  )
```

\newpage

## Exercise 10 

Suppose we run the following chain, **in order**:

```{r}
#| echo: true
#| eval: false
songs |>
  filter(genre != "Indie") |>
  mutate(
    viral_status = if_else(
      streams_millions >= 50,
      "Viral hit",
      "Not yet viral"
    )
  ) |>
  group_by(viral_status) |>
  summarize(count = n())
```


1. Shade the rows below that survive the `filter()` step. (Hint: think about
   what `!=` does with the missing `genre` value, and check your answer with
   `is.na()` logic from Exercise 6 if you're not sure.)

```{r}
#| label: show-final-start
songs |>
  filter(genre != "Indie")
```

2. List which tracks remain after `filter()`: `r (songs[songs$genre != "Indie",])$track_name`

3. Fill in the final summary table:

```{r}
#| label: show-final
songs |>
  filter(genre != "Indie") |>
  mutate(
    viral_status = if_else(
      streams_millions >= 50,
      "Viral hit",
      "Not yet viral"
    )
  ) |>
  group_by(viral_status) |>
  summarize(count = n())
```

\newpage

# Part 2 — Write the Code Yourself

In Part 1, you were given the code and had to work out the output by hand.
Now it's reversed: you'll see the **input** data frame and the **output**
data frame, and your job is to write the dplyr code that turns one into the
other. There can be more than one correct way to write some of these — as
long as your code produces the same output, it's right.

## The Data

This is the data set for Part 2: `movies`, 10 rows and 5 columns.

```{r}
#| label: show-movies-main
movies
```

\newpage

## Exercise 11 

**Output:** (only the column names changed — everything else is identical)

```{r}
#| label: show-ex11-output
movies |>
  rename(movie_genre = genre, box_office_m = box_office_millions)
```

Write the code that produces this output from `movies`.

```{r}
#| label: show-ex11-output2
#| echo: true
#| eval: false
movies |>
  rename(movie_genre = genre, box_office_m = box_office_millions)
```


\newpage

## Exercise 12 

**Output:**

```{r}
#| label: show-ex12-output
movies |>
  select(title, release_year, box_office_millions)
```

Write the code that produces this output from `movies`.

```{r}
#| label: show-ex12-output2
#| echo: true
#| eval: false
movies |>
  select(title, release_year, box_office_millions)
```

\newpage

## Exercise 13 

**Output:**

```{r}
#| label: show-ex13-output
movies |>
  filter(release_year >= 2018)
```

Write the code that produces this output from `movies`.

```{r}
#| label: show-ex13-output2
#| echo: true
#| eval: false
movies |>
  filter(release_year >= 2018)
```


\newpage

## Exercise 14 


**Output:**

```{r}
#| label: show-ex14-output
movies |>
  filter(genre == "Animation" & box_office_millions >= 70)
```

Write the code that produces this output from `movies`.
(Hint: two conditions have to be true at once.)

```{r}
#| label: show-ex14-output2
#| echo: true
#| eval: false
movies |>
  filter(genre == "Animation" & box_office_millions >= 70)
```

\newpage

## Exercise 15 

**Input:**

```{r}
#| label: show-ex15-input
movies
```

**Output:**

```{r}
#| label: show-ex15-output
specific_years <- c(2016, 2019)
movies |>
  filter(release_year %in% specific_years)
```

Write the code that produces this output from `movies`.
(Hint: you'll need a vector of years and the `%in%` operator.)

```{r}
#| label: show-ex15-output2
#| echo: true
#| eval: false
specific_years <- c(2016, 2019)
movies |>
  filter(release_year %in% specific_years)
```

\newpage

## Exercise 16 

**Input:**

```{r}
#| label: show-ex16-input
movies
```

**Output:**

```{r}
#| label: show-ex16-output
movies |>
  filter(is.na(genre))
```

Write the code that produces this output from `movies` while using `is.na()`.

```{r}
#| label: show-ex16-output2
#| echo: true
#| eval: false
movies |>
  filter(is.na(genre))
```

## Exercise 17 

**Output:**

```{r}
#| label: show-ex17-output
movies |>
  select(title, box_office_millions) |>
  mutate(
    hit_status = if_else(
      box_office_millions >= 70,
      "Box-office hit",
      "Modest run"
    )
  )
```

Write the code that produces this output. (You're adding one new column, `hit_status`, based on a condition.)

```{r}
#| label: show-ex17-output2
#| echo: true
#| eval: false
movies |>
  select(title, box_office_millions) |>
  mutate(
    hit_status = if_else(
      box_office_millions >= 70,
      "Box-office hit",
      "Modest run"
    )
  )
```

\newpage

## Exercise 18

**Output:**

```{r}
#| label: show-ex18-output
movies |>
  select(title, box_office_millions) |>
  mutate(
    revenue_tier = case_when(
      box_office_millions < 50 ~ "Under $50M",
      box_office_millions < 80 ~ "$50M to $79M",
      box_office_millions >= 80 ~ "$80M or more"
    )
  )
```

Write the code that produces this output. (You're adding one new column, `revenue_tier`, with three possible values.)

```{r}
#| label: show-ex18-output2
#| echo: true
#| eval: false
movies |>
  select(title, box_office_millions) |>
  mutate(
    revenue_tier = case_when(
      box_office_millions < 50 ~ "Under $50M",
      box_office_millions < 80 ~ "$50M to $79M",
      box_office_millions >= 80 ~ "$80M or more"
    )
  )
```

\newpage

## Exercise 19 


**Output:**

```{r}
#| label: show-ex19-output
movies |>
  group_by(genre) |>
  summarize(
    avg_runtime = mean(runtime_min),
    count = n()
  )
```

Write the code that produces this output. (You're grouping by one column and calculating two summary values per group.)

```{r}
#| label: show-ex19-output2
#| echo: true
#| eval: false
movies |>
  group_by(genre) |>
  summarize(
    avg_runtime = mean(runtime_min),
    count = n()
  )
```


## Exercise 20


**Output:**

```{r}
#| label: show-ex20-output
movies |>
  filter(genre != "Adventure") |>
  mutate(
    hit_status = if_else(
      box_office_millions >= 70,
      "Box-office hit",
      "Modest run"
    )
  ) |>
  group_by(hit_status) |>
  summarize(count = n())
```

Write the full chain of code — filter, then mutate, then
group and summarize — that turns the input into this output. (Hint: check
how many rows disappeared in the `filter()` step, and think about why. Does
`!=` keep or drop a missing `genre` value?)

```{r}
#| label: show-ex20-output2
#| echo: true
#| eval: false
movies |>
  filter(genre != "Adventure") |>
  mutate(
    hit_status = if_else(
      box_office_millions >= 70,
      "Box-office hit",
      "Modest run"
    )
  ) |>
  group_by(hit_status) |>
  summarize(count = n())
```
